Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-07-28 resubmission/blinded-paper-resubmit.tex
Up ajustes (renderizado)

The justification of economic inequality has a long tradition in
sociology, with seminal works such as the ones from Kluegel
(\citeproc{ref-kluegel_social_1995a}{1995};
\citeproc{ref-kluegel_beliefs_1986}{1986}) and Lane
(\citeproc{ref-lane_political_1962}{1962};
\citeproc{ref-lane_market_1986}{1986}). Lane's main aim was to ``help
explain the tenacious hold of market capitalism on the public mind''
(\citeproc{ref-lane_market_1986}{1986, p. 384}), proposing the
distinction between market and political justice. Whereas political
justice refer to government actions aligned with justice principles such
equality or need, market justice involves the allocation linked to
earned desserts, particularly through individual contributions based on
as effort and ability. Sociological survey research in this area has
considered questions related to the extent to which it is considered
just to get better access to social services (such as pensions, health
and education) based on payment capacity
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). Although
these services could be considered basic rights of citizens, evidence to
date suggests that a significant portion of the population justifies
inequality in these areas (\citeproc{ref-koos_moral_2019}{Koos \&
Sachweh, 2019}) and that such preferences vary according to objective
socioeconomic factors and the contextual characteristics of countries
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In the present paper, we
argue that there are two other relevant factors to take into account
when analyzing differences between individuals and groups to this
regard: the perception of economic inequality and the perception of
meritocracy.

@@

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
who makes a contrast between market justice and political justice. The
author defines market justice as a system of ``earned deserts'', whereby
individuals are seen as deserving of a determined distributive outcome
due to their effort and skills. In contrast, political justice
emphasizes principles of equality and need, which are often represented
by the welfare state action through social policies. An important remark
is that the principles of market justice rely on the assumption that
markets are neutral and self-regulating arenas, where individuals are
treated fairly because they face the same formal rules of engagement
(\citeproc{ref-lane_market_1986}{Robert E. Lane, 1986}). Consequently,
the legitimacy of market justice stems from the assumption that
inequalities are not only inevitable but fair---so long as the rules are
transparent and opportunities are open. In this way, market justice
provides a moral justification for inequality by framing it as a
necessary outcome of individual responsibility
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

@@

Regarding the main dependent and independent variables of this study,
Figure~\ref{fig-meanchange} depicts their standardized average values
across survey waves. The results show a notable upward trend in market
justice preferences, particularly in the most recent waves. Perceived
economic inequality exhibited the highest standardized mean in 2016;
however, this variable shows a general downward trajectory over time,
with a temporary increase in 2019, possibly reflecting the effects of
the social uprising that year. Interestingly, while perceptions of
inequality declined in the latest waves (2022--2023), market justice
preferences continued to rise. Meritocratic beliefs, in contrast, remain
relatively stable overall. Nevertheless, perceptions that individuals
are rewarded based on talent tend to be slightly higher than those based
on effort. Both meritocracy-related measures follow a similar temporal
pattern: an increase from 2016 to 2018, a decline between 2019 and
2022---potentially associated with the social unrest and the
consequences of the COVID-19 pandemic---and a subsequent rise in 2023,
returning to levels comparable to those observed in 2016.

@@

When examining the between-group effects (BE) in Model 5 and 6, which
capture differences between individuals in the average of the main
variables, a similar pattern emerges. Individuals who perceive higher
levels of economic inequality tend to prefer less market justice
(\(\beta\) = -0.002, \(p\) \textgreater{} .05). However, this effects is
no longer statistically significant. In Model 6, the meritocratic
perception that effort is rewarded is positively associated with market
justice preferences (\(\beta\) = 0.206, \(p\) \textless{} .001), whereas
the perception that talent is rewarded shows a positive but
non-significant coefficient (\(\beta\) = 0.036).

@@

\caption{\label{tbl-interactions2}Interactions for meritocracy,
perceived economic inequality and market justice preferences}
\scalebox{0.7}{
\begin{tabular}{l c c c c c c}
 & Model 12 & Model 13 & Model 14 & Model 15 & Model 16 & Model 17 \\
Intercept                                                & $1.263^{***}$ & $1.265^{***}$ & $1.246^{***}$ & $1.128^{***}$ & $1.099^{***}$ & $1.090^{***}$ \\
                                                         & $(0.143)$     & $(0.141)$     & $(0.142)$     & $(0.176)$     & $(0.169)$     & $(0.170)$     \\
Wave x Perception inequality (WE)                        & $0.003$       &               &               &               &               &               \\
                                                         & $(0.006)$     &               &               &               &               &               \\
Wave x Merit: effort (WE)                                &               & $0.012$       &               &               &               &               \\
                                                         &               & $(0.006)$     &               &               &               &               \\
Wave x Merit: talent (WE)                                &               &               & $0.007$       &               &               &               \\
                                                         &               &               & $(0.006)$     &               &               &               \\
Wave x Perception inequality (BE)                        &               &               &               & $-0.011$      &               &               \\
                                                         &               &               &               & $(0.009)$     &               &               \\
Wave x Merit: effort (BE)                                &               &               &               &               & $-0.020$      &               \\
                                                         &               &               &               &               & $(0.011)$     &               \\
Wave x Merit: talent (BE)                                &               &               &               &               &               & $-0.016$      \\
                                                         &               &               &               &               &               & $(0.010)$     \\
Controls                                                 & Yes           & Yes           & Yes           & Yes           & Yes           & Yes           \\
BIC                                                      & $31010.853$   & $30888.601$   & $30785.269$   & $31515.978$   & $31493.536$   & $31501.405$   \\
Numb. obs.                                               & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        \\
Num. groups: individuals                                 & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        \\
Var: individuals (Intercept)                             & $0.388$       & $0.334$       & $0.379$       & $0.274$       & $0.532$       & $0.372$       \\
Var: individuals, perception inequality cwc              & $0.100$       & $$            & $$            & $$            & $$            & $$            \\
Var: individuals, wave                                   & $0.025$       & $0.023$       & $0.024$       & $0.021$       & $0.021$       & $0.021$       \\
Cov: individuals (Intercept), perception inequality cwc  & $-0.081$      & $$            & $$            & $$            & $$            & $$            \\
Cov: individuals (Intercept), wave                       & $-0.074$      & $-0.062$      & $-0.072$      & $-0.054$      & $-0.056$      & $-0.049$      \\
Cov: individuals, perception inequality cwc, wave        & $0.021$       & $$            & $$            & $$            & $$            & $$            \\
Var: Residuals                                           & $0.291$       & $0.280$       & $0.275$       & $0.343$       & $0.342$       & $0.343$       \\
Var: individuals, merit effort cwc                       & $$            & $0.122$       & $$            & $$            & $$            & $$            \\
Cov: individuals (Intercept), merit effort cwc           & $$            & $-0.009$      & $$            & $$            & $$            & $$            \\
Cov: individuals, merit effort cwc, wave                 & $$            & $0.000$       & $$            & $$            & $$            & $$            \\
Var: individuals, merit talent cwc                       & $$            & $$            & $0.112$       & $$            & $$            & $$            \\
Cov: individuals (Intercept), merit talent cwc           & $$            & $$            & $0.003$       & $$            & $$            & $$            \\
Cov: individuals, merit talent cwc, wave                 & $$            & $$            & $-0.003$      & $$            & $$            & $$            \\
Var: individuals, perception inequality mean             & $$            & $$            & $$            & $0.000$       & $$            & $$            \\
Cov: individuals (Intercept), perception inequality mean & $$            & $$            & $$            & $0.007$       & $$            & $$            \\
Cov: individuals, perception inequality mean, wave       & $$            & $$            & $$            & $-0.001$      & $$            & $$            \\
Var: individuals, merit effort mean                      & $$            & $$            & $$            & $$            & $0.051$       & $$            \\
Cov: individuals (Intercept), merit effort mean          & $$            & $$            & $$            & $$            & $-0.110$      & $$            \\
Cov: individuals, merit effort mean, wave                & $$            & $$            & $$            & $$            & $-0.001$      & $$            \\
Var: individuals, merit talent mean                      & $$            & $$            & $$            & $$            & $$            & $0.031$       \\
Cov: individuals (Intercept), merit talent mean          & $$            & $$            & $$            & $$            & $$            & $-0.053$      \\
Cov: individuals, merit talent mean, wave                & $$            & $$            & $$            & $$            & $$            & $-0.003$      \\
\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$. CWC = centered within group.}}

With respect to the temporal dynamics of the key predictors,
Table~\ref{tbl-interactions2} shows that time has no statistically
significant effects on either the within- or between-person effects of
perceived inequality and meritocratic perceptions. The within-person
interaction effects are positive, suggesting that in later waves, the
influence of perceived inequality and meritocratic beliefs on market
justice preferences tends to increase. Conversely, in the between-person
specification, the interaction effects are negative, indicating that
over time, the impact of perceived inequality and meritocratic beliefs
on support for market-based justice tends to diminish. However, these
trends are not statistically significant, and thus no definitive
conclusions can be drawn.

@@

This study tests several key hypotheses concerning the relationships
between perceptions of economic inequality, meritocratic perceptions,
and attitudes toward market justice in Chile. It posits that higher
perceived inequality and stronger meritocratic perceptions both
independently contribute to greater support for market-based
distributions of social goods. Furthermore, the study examines how
perceived economic inequality moderates the influence of meritocratic
perceptions on market justice preferences. Additionally, it considers
how major social and political mobilizations, such as those occurring
around 2019 and subsequent years, may have weakened these relationships,
reflecting increased critical perspectives on systemic inequalities and
market mechanisms. The subsequent discussion interprets the empirical
findings in light of these hypotheses, considering their implications
for understanding attitudes toward inequality and social justice in a
context marked by profound social change.

The first hypothesis suggested that perceptions of economic inequality
were strongly linked to attitudes toward market justice. The findings
demonstrated that individuals perceiving higher income disparities
showed stronger support for the distribution of social services based on
individual income, aligning with theories emphasizing that awareness of
inequality could fuel a critical view of such policies
(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). This relationship
reflected a moral economy where perceptions of systemic unfairness
threatened the legitimacy of existing distributions, prompting calls for
greater equity. Interestingly, the negative within-effects observed over
time implied that as perceptions of inequality became more salient, some
individuals experienced a decline in support for market justice,
possibly due to increasing distrust in market mechanisms or
disillusionment with the fairness of the system. This indicated that
while perception heightened awareness, prolonged exposure or social
mobilization could also foster skepticism about whether redistribution
effectively addressed inequality or simply perpetuated systemic flaws.

Regarding the second hypothesis, the results reinforced that
meritocratic perceptions ---particularly those emphasizing individual
effort and talent--- were associated with stronger support for
market-based distribution systems. Individuals who viewed success as
primarily earned through effort were more inclined to justify existing
inequalities, seeing them as outcomes of individual merit rather than
systemic injustice. This aligned with research indicating that
meritocratic narratives served as moral justifications that reinforced
societal stratification (\citeproc{ref-Castillo2012a_justice}{Castillo,
2012}; \citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These
perceptions functioned symbolically to legitimize structural
disparities, reducing support for redistributive measures by framing
inequality as fair and deserved. The findings suggested that these
perceptions bolstered the social hierarchy by fostering acceptance of
inequality as a reflection of individual virtue rather than structural
failure, which was especially pertinent in neoliberal contexts like
Chile's, where market logic heavily influenced social attitudes
(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).

The third hypothesis concerned the moderating role of perceptions of
inequality on the relationship between meritocracy and market justice
attitudes. The analysis indicated that the positive association between
meritocratic perceptions and support for market-based distribution
tended to weaken when perceptions of economic inequality were
particularly high. Although this moderation is not significant for
within-subject effects (i.e.~over time), when taking into account
between-subjects differences we found a positive effect. This appears as
contrary to our initial hypothesis, as it would mean that effect of
meritocratic perceptions becomes stronger as perception of inequality
increases. Nevertheless, we need to be cautious about this
interpretation, as the main effect of inequality perception on market
justice is negative. Therefore, as perceived inequality become less
negative (i.e.~more positive), or closer to zero, then the meritocratic
perception has an stronger positive relationship on market justice
preferences. This could mean that meritocratic perceptions play an
stronger role in justification of market justice inequalities for those
who are less capable or less prone to perceive inequality.

Finally, the fourth hypothesis addressed the impact of recent social
movements, particularly from 2019 onwards, on the relationship between
meritocratic perceptions, perceptions of inequality, and attitudes
toward market justice. The results suggested that these movements
contributed to a more critical stance toward the fairness of existing
distributional mechanisms. Post-mobilization, the positive relationship
between meritocratic perceptions and support for inequality appeared
attenuated, indicating that collective action and public discourse
during this period fostered questioning of the legitimacy of systemic
inequalities justified solely through meritocratic narratives. These
social mobilizations challenged dominant narratives, emphasizing
structural barriers and systemic injustices, which in turn weakened the
association between individual effort and deservingness. This is aligned
with literature emphasizing that social protests act as catalysts for
transforming societal attitudes by disrupting normative beliefs and
encouraging more critical evaluations of institutional legitimacy
(\citeproc{ref-tilly_contentious_2015}{Tilly \& Tarrow, 2015}). The
findings underscored the importance of socio-political context in
shaping and reshaping attitudes toward inequality and justice,
highlighting how collective action could temporarily or permanently
alter the normative foundations that sustained support for market
justice.

@@

This study examined the complex interplay between perceptions of
economic inequality, meritocratic perceptions, and attitudes toward
market justice in Chile from 2016 to 2023, drawing on longitudinal data
from the ELSOC survey. By exploring how subjective assessments and
social contexts influence support for redistribution and market-based
resource allocation, the research offers different elements that
contribute to the understanding of the normative foundations
underpinning social justice attitudes in a highly unequal, neoliberal
environment.

The findings confirm that higher perceptions of economic inequality are
associated with less support for market justice attitudes, this is, the
belief that it is fine that those with higher income have better social
services such as education, pensions and health. Meritocratic
perceptions strongly reinforce support for market justice, serving as
moral justifications for structural inequalities. Additionally,
perceptions of inequality moderate the influence of meritocratic
perceptions, such that as the awareness of disparities are attenuated
(i.e.~less negative) the role of meritocratic perceptions on existing
market-based allocations become stronger. Notably, the social
mobilizations of 2019 and subsequent periods seem to have contributed to
shifting attitudes, fostering greater skepticism of systemic fairness
and reducing the legitimizing power of meritocratic narratives.

This research advances the extant literature by integrating subjective
perceptions with social and political contexts to explain attitudes
toward economic inequality and distributional justice. While previous
studies primarily focused on objective measures or individual
characteristics, this work emphasizes the dynamic and interactional
nature of perceptions and beliefs over time. It highlights the
importance of socio-political upheavals in reshaping normative
attitudes, underscoring the role of collective action in challenging
entrenched narratives of meritocracy and fairness. The longitudinal
approach provides a deeper temporal perspective on how societal events
could influence individual perceptions and preferences.

Despite its contributions, the study has several limitations. First, the
reliance on self-reported perceptions and attitudes may be subject to
social desirability bias. Second, the focus on Chile, while offering
valuable insights into a specific context, limits the generalizability
of results to other countries with different institutional and cultural
backgrounds. Third, the observational nature of the data prevents causal
inferences, and unobserved confounding factors may influence the
observed relationships. Finally, the potential influence of media,
political discourse, and peer effects on perceptions was not directly
assessed. Future research could explore the causal mechanisms linking
perceptions and attitudes, possibly through experimental or mixed-method
designs. Comparative studies across different national contexts would
help determine the universality or specificity of these dynamics.
Additionally, investigating the role of media, political communication,
and education in shaping perceptions of inequality and meritocracy would
deepen understanding of the normative foundations of social justice
attitudes. Finally, examining how these perceptions influence behavioral
outcomes, such as political participation or support for social
movements, would provide valuable insights into the pathways from
beliefs to collective action and policy change.

@@

\bibitem[\citeproctext]{ref-lane_political_1962}
Lane, R. E. (1962). \emph{Political ideology: Why the {American} common
man believes what he does}. Oxford, England: Free Press of Glencoe.

\bibitem[\citeproctext]{ref-lane_market_1986}
Lane, Robert E. (1986). Market {Justice}, {Political Justice}.
\emph{American Political Science Review}, \emph{80}(2), 383--402.
\url{https://doi.org/10.2307/1958264}

@@

\bibitem[\citeproctext]{ref-tilly_contentious_2015}
Tilly, C., \& Tarrow, S. (2015). \emph{Contentious {Politics}} (Second
Edition, New to this Edition:, Second Edition, New to this Edition:).
Oxford, New York: Oxford University Press.

>>>>

The justification of economic inequality has a long tradition in
sociology, with seminal works such as the ones from Kluegel
(\citeproc{ref-kluegel_social_1995a}{1995};
\citeproc{ref-kluegel_beliefs_1986}{1986}) and Lane
(\citeproc{ref-lane_political_1962}{1962},
\citeproc{ref-lane_market_1986}{1986}). Lane's main aim was to ``help
explain the tenacious hold of market capitalism on the public mind''
(\citeproc{ref-lane_market_1986}{1986, p. 384}), proposing the
distinction between market and political justice. Whereas political
justice refer to government actions aligned with justice principles such
equality or need, market justice involves the allocation linked to
earned desserts, particularly through individual contributions based on
as effort and ability. Sociological survey research in this area has
considered questions related to the extent to which it is considered
just to get better access to social services (such as pensions, health
and education) based on payment capacity
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). Although
these services could be considered basic rights of citizens, evidence to
date suggests that a significant portion of the population justifies
inequality in these areas (\citeproc{ref-koos_moral_2019}{Koos \&
Sachweh, 2019}) and that such preferences vary according to objective
socioeconomic factors and the contextual characteristics of countries
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In the present paper, we
argue that there are two other relevant factors to take into account
when analyzing differences between individuals and groups to this
regard: the perception of economic inequality and the perception of
meritocracy.

@@

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
who makes a contrast between market justice and political justice. The
author defines market justice as a system of ``earned deserts'', whereby
individuals are seen as deserving of a determined distributive outcome
due to their effort and skills. In contrast, political justice
emphasizes principles of equality and need, which are often represented
by the welfare state action through social policies. An important remark
is that the principles of market justice rely on the assumption that
markets are neutral and self-regulating arenas, where individuals are
treated fairly because they face the same formal rules of engagement
(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
legitimacy of market justice stems from the assumption that inequalities
are not only inevitable but fair---so long as the rules are transparent
and opportunities are open. In this way, market justice provides a moral
justification for inequality by framing it as a necessary outcome of
individual responsibility
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

@@

Regarding the main dependent and independent variables of this study,
Figure~\ref{fig-meanchange} depicts their standardized average values
across survey waves. The results show a notable upward trend in market
justice preferences, particularly in the most recent waves. Perceived
economic inequality exhibited the highest standardized mean in 2016;
however, this variable shows a general downward trajectory over time,
with a temporary increase in 2019, possibly reflecting the effects of
the social uprising that year. Interestingly, while perceptions of
inequality declined in the latest waves (2022--2023), market justice
preferences continued to rise. Meritocratic perceptions, in contrast,
remain relatively stable overall. Nevertheless, perceptions that
individuals are rewarded based on talent tend to be slightly higher than
those based on effort. Both meritocracy-related measures follow a
similar temporal pattern: an increase from 2016 to 2018, a decline
between 2019 and 2022---potentially associated with the social unrest
and the consequences of the COVID-19 pandemic---and a subsequent rise in
2023, returning to levels comparable to those observed in 2016.

@@

When examining the between-group effects (BE) in Model 5 and 6, which
capture differences between individuals in the average of the main
variables, a similar pattern emerges. Individuals who perceive higher
levels of economic inequality tend to prefer less market justice
(\(\beta\) = -0.002, \(p\) \textgreater{} .05). However, this effect is
no longer statistically significant. In Model 6, the meritocratic
perception that effort is rewarded is positively associated with market
justice preferences (\(\beta\) = 0.206, \(p\) \textless{} .001), whereas
the perception that talent is rewarded shows a positive but
non-significant coefficient (\(\beta\) = 0.036, \(p\) \textgreater{}
.05).

@@

Regarding the temporal dynamics of the key predictors, the analysis
shows that time has no statistically significant effects on the
intrapersonal or interpersonal effects of perceived inequality and
meritocratic perceptions. Further details about this analysis can be
found in supplementary material.

@@

This study tests several key hypotheses concerning the relationships
between perceptions of economic inequality, meritocratic perceptions,
and preferences toward market justice in Chile. It posits that higher
perceived inequality and stronger meritocratic perceptions both
independently contribute to greater support for market-based
distributions of social goods. Furthermore, the study examines how
perceived economic inequality moderates the influence of meritocratic
perceptions on market justice preferences. Additionally, it considers
how major social and political mobilizations, such as those occurring
around 2019 and subsequent years, may have weakened these relationships,
reflecting increased critical perspectives on systemic inequalities and
market mechanisms. The subsequent discussion interprets the empirical
findings in light of these hypotheses, considering their implications
for understanding attitudes toward inequality and social justice in a
context marked by profound social change.

The first hypothesis suggested that perceptions of economic inequality
were strongly linked to preferences for market justice. The findings
demonstrated that individuals perceiving higher income disparities
showed less support for the idea that the access to core social services
depends on individual income, aligning with theories emphasizing that
awareness of inequality could fuel a critical view of such policies
(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). This relationship
reflected a moral economy where perceptions of systemic unfairness
threatened the legitimacy of existing distributions, prompting calls for
greater equity. Similarly, the negative within-effects observed over
time (\(H1_{b}\)) implied that as perceptions of inequality became more
salient, some individuals experienced a decline in support for market
justice, possibly due to increasing distrust in market mechanisms or
disillusionment with the fairness of the system.

Regarding the second hypothesis, the results reinforced that
meritocratic perceptions ---particularly those emphasizing individual
effort--- were associated with stronger support for market-based
distribution systems. Individuals who viewed success as primarily earned
through effort were more inclined to justify existing inequalities,
seeing them as outcomes of individual merit rather than systemic
injustice. This aligned with research indicating that meritocratic
narratives served as moral justifications that reinforced societal
stratification (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These perceptions
functioned symbolically to legitimize structural disparities, reducing
support for redistributive measures by framing inequality as fair and
deserved. The findings suggested that these perceptions bolstered the
social hierarchy by fostering acceptance of inequality as a reflection
of individual virtue rather than structural failure, which was
especially pertinent in neoliberal contexts like Chile's, where market
logic heavily influenced social attitudes
(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).
Interestingly, intraindividual changes in meritocratic perceptions over
time (\(H2_{b}\)) reveal mixed effects. While an increase in the
perception that rewards are based on individual effort is associated
with stronger preferences for market fairness, an increase in the belief
that rewards come from talent is linked to lower support for such
principles. One possible explanation is that effort is generally
perceived as a controllable and malleable trait, which aligns with the
normative foundations of market fairness. Conversely, talent may be
viewed as an innate and less alterable attribute, making talent-based
reward allocation seem less fair or legitimate. Moreover, greater
exposure to real-life situations where outcomes are strongly determined
by inherent characteristics, rather than effort or hard work, may lead
people to question the fairness of market-based reward systems.

The third hypothesis concerned the moderating role of perceptions of
inequality on the relationship between meritocracy and market justice
attitudes. The analysis indicated that the positive association between
meritocratic perceptions and support for market-based distribution
tended to strengthen when perceptions of economic inequality were
particularly high. While this moderation is not significant for
effort-related effects, it is significant for talent in the within- and
between-individual effects. This appears as contrary to our initial
hypothesis, as it would mean that effect of meritocratic perceptions
becomes stronger as perception of inequality increases. Nevertheless, we
need to be cautious about this interpretation, as the main effect of
inequality perception on market justice is negative. Therefore, as
perceived inequality become less negative (i.e.~more positive), or
closer to zero, then the meritocratic perception has an stronger
positive relationship on market justice preferences. This could mean
that meritocratic perceptions play an stronger role in justification of
market justice inequalities for those who are less capable or less prone
to perceive inequality.

Finally, the fourth hypothesis addressed the impact of recent social
movements---particularly those that emerged in 2019---on the
relationship between meritocratic beliefs, perceptions of inequality,
and attitudes toward market justice. The main effect of time suggests
that, on average, support for market justice increased after 2019.
However, the results do not reveal any significant interactions between
distributive mechanisms and market justice attitudes, indicating that
the 2019 social mobilizations did not substantially alter the underlying
associations between these variables. This may suggest a certain
stability or resilience in the normative frameworks that shape
individuals' evaluations of justice, even in the context of major
collective political events. Alternatively, it is possible that while
the protests challenged institutional legitimacy, they did not
fundamentally shift individuals' views regarding how rewards should be
allocated in society.

@@

This study examined the complex interplay between perceptions of
economic inequality, meritocratic perceptions, and preferences toward
market justice in Chile from 2016 to 2023, drawing on longitudinal data
from the ELSOC survey. By exploring how subjective assessments and
social contexts influence support for redistribution and market-based
resource allocation, the research offers different elements that
contribute to the understanding of the normative foundations
underpinning social justice attitudes in a highly unequal and
commoditized environment.

The findings confirm that higher perceptions of economic inequality are
associated with less support for market justice preferences, this is,
the belief that it is fair that those with higher income have better
social services such as education, pensions and health. At the same
time, meritocratic perceptions---particularly those emphasizing
individual effort---are strongly associated with support for
market-based distributions, suggesting that meritocracy serve as a moral
justification for structural inequalities. However, changes in
meritocratic perceptions over time reveal a more nuanced picture: while
increased emphasis on effort reinforces support for market justice,
increased emphasis on talent tends to reduce it. This distinction likely
reflects broader beliefs about the controllability and fairness of
different meritocratic traits. Moreover, the interaction between
inequality perceptions and meritocratic perceptions indicates that the
legitimizing power of meritocracy becomes stronger as perceptions of
inequality become less negative---that is, when individuals are less
aware or concerned about economic disparities. This finding highlights a
potential feedback mechanism by which lower sensitivity to inequality
may enable stronger endorsement of merit-based explanations for unequal
outcomes.

This research advances the extant literature by integrating subjective
perceptions with social and political contexts to explain attitudes
toward economic inequality and distributional justice. While previous
studies primarily focused on objective measures or individual
characteristics (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}), this work emphasizes the
dynamic and interactional nature of perceptions and beliefs over time.
It highlights the importance of socio-political upheavals in reshaping
normative attitudes, underscoring the role of collective action in
challenging entrenched narratives of meritocracy and fairness. The
longitudinal approach provides a deeper temporal perspective on how
societal events could influence individual perceptions and preferences.

Despite its contributions, the study has several limitations. First, the
reliance on self-reported perceptions and attitudes may be subject to
social desirability bias. Second, the focus on Chile, while offering
valuable insights into a specific context, limits the generalizability
of results to other countries with different institutional and cultural
backgrounds. Third, the observational nature of the data prevents causal
inferences, and unobserved confounding factors may influence the
observed relationships. Finally, the potential influence of media,
political discourse, and peer effects on perceptions was not directly
assessed.

Future research could explore the causal mechanisms linking perceptions
and attitudes, possibly through experimental or mixed-method designs.
Comparative studies across different national contexts would help
determine the universality or specificity of these dynamics.
Additionally, investigating the role of media, political communication,
and education in shaping perceptions of inequality and meritocracy would
deepen understanding of the normative foundations of social justice
attitudes. Finally, examining how these perceptions influence behavioral
outcomes, such as political participation or support for social
movements, would provide valuable insights into the pathways from
beliefs to collective action and policy change.

@@

\bibitem[\citeproctext]{ref-lane_political_1962}
Lane, R. (1962). \emph{Political ideology: Why the {American} common man
believes what he does}. Oxford, England: Free Press of Glencoe.

\bibitem[\citeproctext]{ref-lane_market_1986}
Lane, R. (1986). Market {Justice}, {Political Justice}. \emph{American
Political Science Review}, \emph{80}(2), 383--402.
\url{https://doi.org/10.2307/1958264}

@@
